---
title: "Database connections in R"
subtitle: "Session 11"
institute: "NHS-R Community"
output:
  xaringan::moon_reader:
    css: 
      - default
      - css/nhsr.css
      - css/nhsr-fonts.css
    lib_dir: libs
    seal: false
    self_contained: true
    nature:
      highlightStyle: googlecode
      highlightLines: true
      highlightLanguage: ["r"]
      countIncrementalSlides: false
      ratio: "16:9"
    includes:
      after_body: [css/insert-logo.html]
---

```{r libs, include=FALSE}
library(knitr)
library(tidyverse)
library(xaringan)
library(magick)
library(dplyr)
library(dbplyr)
library(icons)
library(xaringanExtra)
library(RSQLite)

xaringanExtra::use_share_again() # need to get the slide button on html view

opts_chunk$set(
  echo = FALSE,
  eval = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7.252,
  fig.height = 4,
  dpi = 300,
  dev.args = list(type = "cairo")
)

#############################################################################
# Set up exercise database.  This will be held in memory, using RSQLite
# You would likely be connecting to SQL Server, Oracle or similar in reality
#############################################################################

# Load beds_data.csv
beds <- read_csv("https://raw.githubusercontent.com/nhs-r-community/intro_r_data/26a2e39bd49d4aa95a8efc5b45c398386e6e4ed4/beds_data.csv", 
                      col_types = cols(date = col_date(format = "%d/%m/%Y")), 
                      skip = 3)

beds$year <- as.double(substring(beds$date,1,4))
beds$month <- as.double(substring(beds$date,6,7))
beds$day <- as.double(substring(beds$date,9,10))

# Set up the database connection
DBI_connection <- dbConnect(RSQLite::SQLite(), ":memory:")

# Write beds table to the database
dbWriteTable(DBI_connection, "MyTable", beds)

#Remove the imported data frame so we have to work on the database
rm(beds)

############################################################################

```




class: title-slide, left, bottom

# `r rmarkdown::metadata$title`
----
## **`r rmarkdown::metadata$subtitle`**
### `r rmarkdown::metadata$author`
### `r rmarkdown::metadata$date`

---
  
# Two common methods
  
There are two common methods of connection, both of which use Open Database Connectivity (ODBC) drivers:
  
1.  The `RODBC` package (older method, not shown here)
2.  The `DBI` system, using `dplyr,` `dbplyr` and `odbc`.

<br>
+ Both of these create a connection, using a 'connection string' with server details.
+ This can be used to create a connection object that you use in `R`.
+ We can use the object to pull data into `R` or manipulate it in the database.

--

<br>

For these examples, we will assume:

+ SQL Server connection, call 'MyServer'
+ A database called 'MyDB'
+ A table called 'MyTable'

---

# DBI connection

Requires a couple of packages, and knowing your connection details:
+ `DBI` - a common Database Interface engine for use in `S` and `R` (see <a href="https://cran.r-project.org/web/packages/DBI/vignettes/DBI-1.html">here</a>)
+ `odbc`- provides the odbc drivers, but you could use the functions below with other drivers instead.

```{r DBI1, message=FALSE, warning=FALSE, echo=TRUE}
library(DBI)
library(odbc)

DBI_connection <- dbConnect(odbc()
                     , Driver="SQL Server"
                     , Server="MyServer"
                     , database = "MyDB"
                     , trusted_connection="true" 
                     
)
```

--

+ `trusted_connection=true` passes your windows credentials to the server.  
  + Sometimes this is: `Trusted_Connection` and `yes` or `no`, instead of `true` or `false`.
+ You can, instead, specify a username (`uid`) and a password (`pwd`)


---

# Using SQL & returning data to `R` (1)
  
+ Can write an SQL query directly using the `dbSendQuery` function.
+ Executes the query on the server-side only.
+ If you want the results back in `R`, you need to use `dbFetch` as well.

```{r DBIQuery, warning=FALSE, message=FALSE, echo=TRUE}
SomeRecords <- 
  dbFetch(
    dbSendQuery(DBI_connection
                , "Select TOP 100 * from MyTable"
                )
    )

#or

SomeRecords <- 
  dbSendQuery(DBI_connection
              , "Select TOP 100 * from MyTable") %>%
  dbFetch()
```

---

# Using tables in the database
  
Now we can define a table as if it was part of our `R` workspace, using the connection object and the names of the table in the database. 
+ `dplyr` - to make the `tbl` and use it, we'll work with `dplyr` syntax.
+ `dbplyr` - this add-on package allows translation from `dplyr` to SQL.

--

```{r dplyr1, message=FALSE, warning=FALSE, collapse=TRUE, eval=TRUE, echo=TRUE}
MyTable <- tbl(DBI_connection, "MyTable")
```
```{r dplyr2, message=FALSE, warning=FALSE, collapse=TRUE, echo=TRUE}
# Or if your table was in a schema called 'ref', rather than default like'dbo'
MyTable_inschema <- tbl(DBI_connection, in_schema("ref", "MyTable"))
```

--

`glimpse` is a useful function that shows you a summary:
```{r dbplyr5, eval=TRUE, eval=TRUE, echo=TRUE}
glimpse(MyTable)
```


---

# Constructing `dplyr` query

+ We can then perform select or aggregate queries without translation.
+ Even though it returns results, the data are still in the database


```{r dplyr5, message=FALSE, warning=FALSE, collapse=FALSE, eval=TRUE, echo=TRUE}
library(dplyr)
library(dbplyr)

MyTable %>%
  filter(year == 2015) %>%
  group_by(org_name) %>%
  summarise(AvgEvents = mean(beds_av)) %>%
  arrange(month)
```


---

# Returning data to `R` with `dbplyr`

+ May need to pull the data from the server into memory in `R` sometimes.
+ Can do this with `collect()`

```{r dplyrCollect, warning=FALSE, message=FALSE,  eval=TRUE, echo=TRUE}
MyTable_local<- MyTable %>%
  filter(year == 2015) %>%
  group_by(org_name) %>%
  summarise(AvgEvents = mean(beds_av)) %>%
  arrange(org_name) %>%
  collect()

MyTable_local
```

---
  
# Writing to databases
  
You can also write back to a database using the `dbWriteTable` function.
+ For example:
+ Writing a new table current connection, call the table: 'NewDatabaseTable'
+ We are writing back the `R` `data.frame` called "MyTable_local" (that we created in the last section)
+ First time, `dbWriteTable` will create the table.
+ Subsequent uses, you can specify whether to `append` or `overwrite` options using TRUE/FALSE.

```{r dplyrWrite, warning=FALSE, message=FALSE, eval=FALSE, echo=TRUE}
dbWriteTable(DBI_connection
             ,"NewDatabaseTable"
             , MyTable_local
             , overwrite=TRUE)
```

---
class: middle, center

# Exercise:
### Follow the instructions in the `11_database_excerise.R` file

---

# Useful DBI commands

|Command        | Summary                                                |
| --------------|:------------------------------------------------------:|
|dbConnect()    | Create a DBI connection object                         |
|dbListTables() | List the tables on the connection                      |
|dbListFields() | List the fields for a given table on a given connection|
|dbSendQuery()  | Send a query to execute on the server/connection      |
|dbFetch()      | Fetch the results from the server/connection           |
|dbWriteTable() | Write a table to the connection                        |
|               |                                                        |
|tbl()          | Set a table on the connection as a 'tibble' for `dplyr`|
|glimpse()      | See a summary of the rows, data types and top rows     |
|in_schema()    | Use a named schema, other than 'dbo' (`dbplyr` package)  |

---

#### This work is licensed as
</br> Creative Commons
</br> Attribution
</br> ShareAlike 4.0
</br> International
</br> To view a copy of this license, visit
</br> https://creativecommons.org/licenses/by/4.0/
